Likelihood-Based Inference for Generalized Linear Mixed Models: Inference with Package \pkg{glmm}

### 2017-02-14 JSS/article.tex
edited section 2
rearranged, adding subsection headers, removed equation-y stuff for the most part. opted to describe the importance sampling distribution and refer to my thesis rather than describe it in detail in this paper.

@@

Let $\tilde{f}(u)$ be a nonnegative function whose support contains the support of the $f_\theta(u,y)$, where $\tilde{f}(u_k)$ does not depend on $\theta$. Let\hbox{ $u_k,$ $k=1,$ $\ldots,$ $m$}, be vectors of length $q$ drawn from  $\tilde{f}(u_k)$.   The specific importance sampling distribution used in \proglang{R} package \texttt{glmm} is described in Equation \ref{eq:ftwiddle}. 
Then the Monte Carlo  likelihood approximation is
\begin{align}
L_{m}(\theta|y) &=   \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}_. \label{eq:MCLAval}
\end{align}
Again, the function in Equation \ref{eq:MCLAval} approximates the entire likelihood function and can be used for any likelihood-based inference.\\

The asymptotic properties of the likelihood approximation as the Monte Carlo sample size increases have been studied in generality  \citep{geyer:1994, geyer:thom:1992} and   for the special case of an importance sampling distribution chosen independently of the data \citep{sung:geyer:2007}.  \citet{geyer:1994} presents conditions under which  the Monte Carlo likelihood approximation converges almost surely to the exact likelihood for any single parameter value.  In addition to studying the point-wise convergence,  we can also consider  the convergence of the entire likelihood function.
\citet{geyer:1994} shows the Monte Carlo likelihood approximation converges almost surely to the likelihood function for GLMMs specified with  unnormalized densities as long as a Wald-like integrability condition is met.
 Additionally, if the parameter space can be compactified, the MCMLE converges to the MLE almost surely \citep{geyer:1994}. \citet{geyer:1994} also shows that the Monte Carlo profile likelihoods converge to the exact profile likelihoods almost surely, and no additional regularity conditions are required.\\

Prior to \pkg{glmm}, MCLA for GLMMs has  been studied by \citet{sung:geyer:2007}, who  focus on an MCLA implementation with an importance sampling distribution constructed independently of the observed data. \citet{sung:geyer:2007} prepared the \proglang{R} package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data. Though their package can perform maximum likelihood for simpler models,  it cannot find MLEs for a model as complicated as the salamander model (described in section \ref{sec:examples}).  \\

This highlights the challenge of  finding an importance sampling distribution that performs well in practice. Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require so much computing power that modern computers cannot perform maximum likelihood in practice.
  This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems. In response, we present an importance sampling distribution (Equation  \ref{eq:ftwiddle}) and the \proglang{R} package \pkg{glmm}.\\

\subsection[Implementation of MCLA in glmm]{Implementation of MCLA in  \pkg{glmm}}

 and mention that we first need to do PQL. Also Sigma.

Let $D$ be the variance-covariance matrix for the random effect vector $u$. Though $D$ is a function of the variance components, we suppress this notation for simplicity.
Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is, $s$ is defined such that $u= D^{1/2}s$. Let $\sigma$ be a vector of length $K$ with components $\sqrt{\nu_t},$ $t=1,$ $\ldots,$ $K$.  Let $\beta^*$, $s^*$ and $\sigma^*$ denote the PQL estimates for $\beta$, $s$ and $\sigma$, respectively.  Let
\begin{align}
A^*=\sum_{t=1}^K E_t \, \sigma^*_t \nonumber
\end{align}\
and
\begin{align}
 D^* =A^*A^*. \nonumber
\end{align}
 We can ``unstandardize'' our PQL-predicted random effects:
\begin{align}
u^*&=A^*s^*. \nonumber
\end{align}

Let $p_1,p_2,p_3$ be  proportions  such that  \hbox{$p_1+p_2+p_3=1$}.  Let $f(u|\mu,\Sigma)$ denote the density for a multivariate normal distribution with mean $\mu$ and variance matrix $\Sigma$.   
Let $\grave{f}(u|0,D^*)$ denote the density for  a $q$-dimensional multivariate  $t$ distribution with mean $0$, scale matrix $D^*$, and $\zeta=5$ degrees of freedom. Then we propose the following importance sampling distribution:
\begin{align}\label{eq:ftwiddle}
\begin{split}
 \tilde{f}(u) &= p_1  \grave{f}(u|0,D^*)+p_2  f(u \, | \, u^*, D^*)+\\
&\qquad + p_3  f(u \, | \, u^*, (Z^T  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). 
\end{split}
\end{align}

The first component  $\grave{f}( \,  \cdot \, | 0, D^*)$ of Equation \ref{eq:ftwiddle}  is chosen to ensure  the gradient of the MCLA has a central limit theorem \citep{mythesis}.  The second component is chosen because it is centered at  $u^*$ and  has  variance $D^*$. The last component is centered at $u^*$ and has a variance based on the Hessian of the  penalized likelihood from PQL.  More specifically, the Hessian of the log density of the last distribution matches the Hessian of the log density of the target distribution $f_\theta(u,y)$.

>>>>

Prior to \pkg{glmm}, MCLA for GLMMs was   studied by \citet{sung:geyer:2007}, who  focused on an MCLA implementation with an importance sampling distribution constructed independently of the observed data. \citet{sung:geyer:2007} prepared the \proglang{R} package \texttt{bernor} that  fits maximum likelihood estimates for a GLMM with a Bernoulli response; their package uses an importance sampling distribution that does not depend on the observed data. Though their package can perform maximum likelihood for simpler models,  it cannot find MLEs for a model as complicated as the salamander model (described in section \ref{sec:examples}).  \\

This highlights the challenge of  finding an importance sampling distribution that performs well in practice. Though MCLA theory indicates any importance sampling distribution should suffice as long as its support contains the support of the target distribution, many importance sampling distributions require so much computing power that modern computers cannot perform  likelihood-based inference in practice.
  This practical problem is  illustrated by the absence of MCLA software that can produce MLEs for nontrivial problems. In response, we present an importance sampling distribution \citep{mythesis} and the \proglang{R} package \pkg{glmm}.\\

\subsection{Calculation of the Monte Carlo likelihood approximation}

@@

Let $\tilde{f}(u)$ be a nonnegative function whose support contains the support of the $f_\theta(u,y)$, where $\tilde{f}(u_k)$ does not depend on $\theta$.  The  importance sampling distribution used in \proglang{R} package \texttt{glmm} is described in section \ref{sec:myimplem} and by \citet{mythesis}.  Let\hbox{ $u_k,$ $k=1,$ $\ldots,$ $m$}, be vectors of length $q$ drawn from  $\tilde{f}(u_k)$.  
Then the Monte Carlo  likelihood approximation is
\begin{align}
L_{m}(\theta|y) &=   \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(u_k,y)   }{\tilde{f}(u_k)}_. \label{eq:MCLAval}
\end{align}
Because the function in Equation \ref{eq:MCLAval} approximates the entire likelihood function, it can be used to conduct any likelihood-based inference.\\

\subsection[Implementation of MCLA in glmm]{Implementation of MCLA in  \pkg{glmm}} \label{sec:myimplem}

The importance sampling distribution used in \pkg{glmm} depends on the observed data and is specified in detail by \citet{mythesis}. It is a mixture of three distributions: two normal distributions and a t distribution. The t distribution is centered at 0 with a covariance matrix that depends on penalized quasi-likelihood (PQL) estimates of the variance components. The first normal distribution is centered at the PQL predictions of the random effects and has a variance matrix that depends on the PQL estimates of the variance components. The second normal distribution is  centered at the PQL predictions of the random effects and has a covariance matrix based on the Hessian of the  penalized likelihood from PQL.  More specifically, the Hessian of the log density of the last distribution matches the Hessian of the log density of the target distribution $f_\theta(u,y)$.

\subsection{Asymptotic properties of the Monte Carlo likelihood approximation}

The asymptotic properties of the likelihood approximation as the Monte Carlo sample size increases have been studied in generality  \citep{geyer:1994, geyer:thom:1992} and   for the specific implementation in \pkg{glmm} \citep{mythesis}.  \citet{geyer:1994} presents conditions under which  the Monte Carlo likelihood approximation converges almost surely to the exact likelihood for any single parameter value.  In addition to studying the point-wise convergence,  we can also consider  the convergence of the entire likelihood function.
\citet{geyer:1994} shows the Monte Carlo likelihood approximation converges almost surely to the likelihood function for GLMMs specified with  unnormalized densities as long as a Wald-like integrability condition is met.
 Additionally, if the parameter space can be compactified, the MCMLE converges to the MLE almost surely \citep{geyer:1994}. \citet{geyer:1994} also shows that the Monte Carlo profile likelihoods converge to the exact profile likelihoods almost surely, and no additional regularity conditions are required.\\
